Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2022-05-16 2022/eamt/eamt-latex/eamt22.tex
local changes

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key aspect of training is then to mitigate the detrimental effects of a mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from; this setting differs from multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \cite{Chu2017comparison,Saunders21asurvey} for a review), and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \cite{Shen21source}} -> I think this is not so appropriate here
where the diversity of training data not only corresponds to variations in the topic, genre, or register but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the testing conditions \cite{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however, a challenging task due, for instance, to the similarity between domains/languages, or to the regularization effects of out-of-domain data \cite{Miceli-barone17regularization}. A static policy may also be suboptimal when some target domains or languages are easier to train than others. Finally, improving the performance of the MT system in one domain will often hurt that of another \cite{Vanderwees17dynamic,Britz17mixing} and improving model generalization across all domains \cite{koehn18findings} may not achieve optimally for any particular domain.

@@

We conventionally define a domain $d$ as a distribution $\mathcal{D}_d(x)$ over some feature space $\mathcal{X}$ that is shared across domains \cite{Pan10asurvey}: in machine translation, $\mathcal{X}$ is the representation space for input sentences; each domain corresponds to a specific source of data, and may differ from other data sources in terms of textual genre, thematic content \cite{Chen16guided,Zhang16topicinformed}, register, style \cite{Niu18multitask}, etc. Translation in domain $d$ is formalized by a translation function $h_d(y|x)$ pairing sentences in a source language with sentences in a target language $y \in \mathcal{Y}$. $h_d$ is usually assumed to be deterministic (hence $y = h_d(x)$) but may differ across domains.

@@

These situations have been amply documented from a theoretical perspective (eg.\ \cite{Mansour09multiple,Mansour09domainadaptation,Hoffman18algorithms}). A general recommendation in the DA setting is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

@@

We experiment with translation from English into French in 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}); the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Additional experiments use the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website\footnote{\url{http://opus.nlpl.eu}}. These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu} \cite{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

Domain adaptation is an old problem that has been studied from many angles, both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \cite{Chu2017comparison}, where the authors distinguish between data-centric and model-centric DA, a view also adopted in the recent survey of \newcite{Saunders21asurvey}. Our approach to DA in this paper falls under the former category. We refer readers interested in DA to these papers.

Multidomain NMT (MDMT) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDMT combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \cite{Jiang19multidomain}. Unsupervised MDMT is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive/dynamic data selection take inspiration from \newcite{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of the ``easiness'' of a sample to schedule the presentation of training data so as to start with the easiest examples and end with the hardest. Various ways to automate CL in the framework of multi-armed bandits are explored in \cite{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for multidomain and multilingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \newcite{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: early stages only use in-data, while less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is opposite to the policy of \newcite{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively increases the focus on in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\newcite{Kumar19reinforcement} use reinforcement learning to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the devset loss that results from the current data selection strategy.\fyDone{Alert: what do we do during warm-up ?} This technique is applied to multilingual NMT in \cite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another CL-based approach which relies on \emph{instance uncertainty} as a measure of their difficulty and presents data samples starting with the easiest. Another contribution of this work is a new stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multidomain NMT, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied to train a sequence of models. The one that maximizes the devset performance is finally retained.

>>>>

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key aspect of training is then to mitigate the detrimental effects of a mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from; this setting differs from multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \newcite{Chu2017comparison} and \newcite{Saunders21asurvey} for a review), and so is multi-domain (MD) learning \cite{Chu18multilingual,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \cite{Shen21source}} -> I think this is not so appropriate here
where the diversity of training data not only corresponds to variations in the topic, genre, or register but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the testing conditions \cite{Axelrod11domain}. Finding the optimal balance of training data is however, a challenging task due, for instance, to the similarity between domains/languages, or to the regularization effects of out-of-domain data \cite{Miceli-barone17regularization}. A static policy may also be suboptimal when some target domains or languages are easier to train than others. Finally, improving the performance of the MT system in one domain will often hurt that of another \cite{Vanderwees17dynamic,Britz17mixing} and improving model generalization across all domains \cite{koehn18findings} may not achieve optimally for any particular domain.

@@

We conventionally define a domain $d$ as a distribution $\mathcal{D}_d(x)$ over some feature space $\mathcal{X}$ that is shared across domains \cite{Pan10asurvey}: in machine translation, $\mathcal{X}$ is the representation space for input sentences; each domain corresponds to a specific source of data, and may differ from other data sources in terms of textual genre, thematic content \cite{Chen16guided}, register, style \cite{Niu18multitask}, etc. Translation in domain $d$ is formalized by a translation function $h_d(y|x)$ pairing sentences in a source language with sentences in a target language $y \in \mathcal{Y}$. $h_d$ is usually assumed to be deterministic (hence $y = h_d(x)$) but may differ across domains.

@@

These situations have been amply documented from a theoretical perspective \ \cite{Mansour09multiple,Mansour09domainadaptation,Hoffman18algorithms}. A general recommendation in the DA setting is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

@@

We experiment with translation from English into French in 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}); the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Additional experiments use the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website\footnote{\url{http://opus.nlpl.eu}}. These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Post18call}.\footnote{We use truecasing and \texttt{sacrebleu} \cite{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

Domain adaptation is an old problem that has been studied from many angles, both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \newcite{Chu2017comparison}, where the authors distinguish between data-centric and model-centric DA, a view also adopted in the recent survey of \newcite{Saunders21asurvey}. Our approach to DA in this paper falls under the former category. We refer readers interested in DA to these papers.

Multidomain NMT (MDMT) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDMT combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing}; (c) the use of automatic domain identification \cite{Jiang19multidomain}. Unsupervised MDMT is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive/dynamic data selection take inspiration from \newcite{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of the ``easiness'' of a sample to schedule the presentation of training data so as to start with the easiest examples and end with the hardest. Various ways to automate CL in the framework of multi-armed bandits are explored in \newcite{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for multidomain and multilingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \newcite{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: early stages only use in-data, while less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is opposite to the policy of \newcite{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively increases the focus on in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\newcite{Kumar19reinforcement} use reinforcement learning to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the devset loss that results from the current data selection strategy.\fyDone{Alert: what do we do during warm-up ?} This technique is applied to multilingual NMT in \newcite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another CL-based approach which relies on \emph{instance uncertainty} as a measure of their difficulty and presents data samples starting with the easiest. Another contribution of this work is a new stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multidomain NMT, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied to train a sequence of models. The one that maximizes the devset performance is finally retained.
